# Chapter 6 — Manual chill analysis

In this chapter, we calculate warm hours above 25°C using the `chillR` package.


## Short code

### Exercise 1 & 2
**Write a basic function that calculates warm hours (>25°C) & Apply this function to the Winters_hours_gaps dataset**


```{r}
library(chillR)
data(Winters_hours_gaps)

hours_over_25 <- sum(Winters_hours_gaps$Temp > 25, na.rm = TRUE)
print(paste("Total hours over 25°C:", hours_over_25))

```

### Exercise 3 
**Extend this function, so that it can take start and end dates as inputs and sums up warm hours between these dates**

```{r}
calculate_warm_hours <- function(data, start_date, end_date) {
  data$Date <- as.Date(paste(data$Year, data$Month, data$Day, sep = "-"))

  data_filtered <- data[data$Date >= as.Date(start_date) & data$Date <= as.Date(end_date), ]

  warm_hours <- sum(data_filtered$Temp > 25, na.rm = TRUE)
  return(warm_hours)
}

start <- "2008-06-01"
end <- "2008-09-28"

result <- calculate_warm_hours(Winters_hours_gaps, start, end)
print(paste("Warm hours between", start, "and", end, ":", result))
```

## Longer code

### Exercise 1
**Write a basic function that calculates warm hours (>25°C)**

```{r}
library(chillR)
data(Winters_hours_gaps)
```


```{r}
hours_over_25 <- sum(Winters_hours_gaps$Temp > 25, na.rm = TRUE)
print(paste("Total hours over 25 degrees:", hours_over_25))
```

### Exercise 2 
**Apply this function to the Winters_hours_gaps dataset**

I applied this calculation directly to the dataset `Winters_hours_gaps`, giving a total of **1275 warm hours** throughout the recorded period.

This means there were 1275 hours in 2008 where the temperature went above 25°C.

### Exercise 3 
**Extend this function, so that it can take start and end dates as inputs and sums up warm hours between these dates**

I extended the code to define a reusable function that calculates warm hours between **two specific dates**.

```{r}
calculate_warm_hours_final_filter <- function(data, start_date_char, end_date_char) {
  start_date_obj <- as.Date(start_date_char)
  end_date_obj <- as.Date(end_date_char)

  data$Date <- as.Date(paste(data$Year, data$Month, data$Day, sep = "-"))

  filtered_data <- data[data$Date >= start_date_obj & data$Date <= end_date_obj, ]

  print(paste("Number of hourly records in filtered data:", nrow(filtered_data)))

  hours_over_25 <- sum(filtered_data$Temp > 25, na.rm = TRUE)

  return(hours_over_25)
}

start_date_input_user <- "2008-06-01"
end_date_input_user <- "2008-09-28"

total_warm_hours_user <- calculate_warm_hours_final_filter(
  Winters_hours_gaps,
  start_date_input_user,
  end_date_input_user
)

print(paste(
  "Total hours over 25 degrees between",
  start_date_input_user, "and", end_date_input_user, ":",
  total_warm_hours_user
))


```
