# Chapter 6 — Manual chill analysis

In this chapter, we calculate warm hours above 25°C using the `chillR` package.


## Calculating warm hours (>25 °C)

**Write a basic function that calculates warm hours (>25°C) & Apply this function to the Winters_hours_gaps dataset.**

```{r}
library(chillR)
data(Winters_hours_gaps)

hours_over_25 <- sum(Winters_hours_gaps$Temp > 25, na.rm = TRUE)
print(paste("Total hours over 25°C:", hours_over_25))

```

## Calculating warm hours for a defined time period

**Extend this function, so that it can take start and end dates as inputs and sums up warm hours between these dates**

```{r}
calculate_warm_hours <- function(data, start_date, end_date) {
  data$Date <- as.Date(paste(data$Year, data$Month, data$Day, sep = "-"))

  data_filtered <- data[data$Date >= as.Date(start_date) & data$Date <= as.Date(end_date), ]

  warm_hours <- sum(data_filtered$Temp > 25, na.rm = TRUE)
  return(warm_hours)
}

start <- "2008-06-01"
end <- "2008-09-28"

result <- calculate_warm_hours(Winters_hours_gaps, start, end)
print(paste("Warm hours between", start, "and", end, ":", result))
```

