# Chapter 7 — Chill models

In this chapter, we explore how chill models quantify winter cold accumulation and learn to run, customize, and apply these models to temperature data.


## Applying a built-in chill model

**Run the chilling function on the Winters_hours_gap dataset**

```{r}

library(chillR)
data("Winters_hours_gaps")
output_chilling <- chilling(make_JDay(Winters_hours_gaps),
                            Start_JDay = 90,  
                            End_JDay = 100)
print(output_chilling)

```

## Defining a custom chill model

**Create your own temperature-weighting chill model using the step_model function**

```{r}

my_chill_model <- data.frame(
  lower = c(-1000, 0, 3, 6, 9, 12, 15),
  upper = c(0, 3, 6, 9, 12, 15, 1000),
  weight = c(0, 0.5, 1, 0.5, 0, -0.5, -1)
)

print(my_chill_model)

my_custom_chill <- function(x) step_model(x, my_chill_model)

my_custom_chill(Winters_hours_gaps$Temp)[1:20]

```

## Applying the custom chill model

**Run this model on the Winters_hours_gaps dataset using the tempResponse function.**

```{r}

output_custom <- tempResponse(make_JDay(Winters_hours_gaps),
                              Start_JDay = 90,
                              End_JDay = 100,
                              models = list(MyChill = my_custom_chill))

print(output_custom)

```
