# Chapter 8 — Making hourly temperatures

In this chapter, we will explore how to create hourly temperature data from daily records using the `chillR` package. We will use both idealized and empirical temperature curves.

## Daylength and solar geometry

**Choose a location of interest, find out its latitude and produce plots of daily sunrise, sunset and daylength**

Here, I use **Edinburgh, Scotland (latitude = 55.95)** as the example location.

```{r}
library(chillR)
library(ggplot2)
library(tidyr)
latitude <- 55.95
Days <- chillR::daylength(latitude = 55.95, JDay = 1:365)
Days_df <-
  data.frame(
    JDay = 1:365,
    Sunrise = Days$Sunrise,
    Sunset = Days$Sunset,
    Daylength = Days$Daylength
  )
Days_df <- pivot_longer(Days_df,cols=c(Sunrise:Daylength))

ggplot(Days_df, aes(JDay, value)) +
  geom_line(lwd = 1.5, col = "darkgreen") +
  facet_grid(cols = vars(name)) +
  ylab("Time of Day / Daylength (Hours)") +
  xlab("Julian Day") +
  ggtitle("Sunrise, Sunset and Daylength Over the Year") +
  theme_bw(base_size = 16)
```


## Idealized diurnal temperature curves

**Produce an hourly dataset, based on idealized daily curves, for the KA_weather dataset**

```{r}
latitude <- 55.95
KA_hourly <- stack_hourly_temps(KA_weather, latitude = latitude)$hourtemps
head(KA_hourly, 24)
```

## Empirical reconstruction of hourly temperatures

**Produce empirical temperature curve parameters for the Winters_hours_gaps dataset, and use them to predict hourly values from daily temperatures.**

Sometimes the idealized curve may not reflect real conditions (e.g., topographic or local effects).
In such cases, empirical curves are better. They are based on real hourly observations.

```{r}
empi_curve <- Empirical_daily_temperature_curve(Winters_hours_gaps)
ggplot(data = empi_curve[1:96, ], aes(Hour, Prediction_coefficient)) +
  geom_line(lwd = 1.3, col = "red") +
  facet_grid(rows = vars(Month)) +
  xlab("Hour of the Day") +
  ylab("Prediction Coefficient") +
  theme_bw(base_size = 10)
Winters_daily <- make_all_day_table(Winters_hours_gaps, input_timestep = "hour")
Winters_emp_hourly <- Empirical_hourly_temperatures(Winters_daily, empi_curve)
head(Winters_emp_hourly, 24)

```
