# Chapter 10 — Getting temperature data

In this chapter, we learn how to obtain and prepare historical temperature data for a chosen location, using the Lake Constance region as an example.


## Selecting a study location and nearby weather stations

**Choose a location of interest and find the 25 closest weather stations using the handle_gsod function**

**Choice of study region:**

- In general, I wanted to use the Lake     Constance region because it is a major   fruit production area in Germany
- Initially, I planned to use the same     location, Gottmadingen                   (Baden-Württemberg), as in my            bachelor’s thesis
- I quickly realized that there is no      nearby weather station for               Gottmadingen
- I then considered Consatnce, but the     available weather data only extend       until 2003
- Another option was the weather station   in Schaffhausen, which has good data     coverage
- However, I found the milder Lake         Constance climate more interesting for   the analysis
- The next weather station with nearly     complete data is Güttingen (Thurgau),    Switzerland, which is also located       directly on Lake Constance

**Climatic and agricultural relevance of the region:**

- The Lake Constance region typically has   mild, warm autumns, which are very       favorable for fruit production and       especially viticulture
- Large day–night temperature differences   are a key factor that support optimal    fruit ripening
- Güttingen hosts an agricultural          experimental farm, the Schul- und        Versuchsbetrieb in Güttingen, which is   actively used for research in fruit and   berry cultivation
- At this site, more than 10 fruit and     berry species and over 100 different     varieties are cultivated
- The Swiss canton of Thurgau is the       largest fruit-growing region in          Switzerland, with a gross income of      approximately 80 million Swiss francs    from fruit production
- Like most regions, Thurgau is affected   by climate change, with meteorological   records dating back to 1864 showing      clear signals of climatic change
- Extreme weather events have become more   frequent, including heatwaves (e.g.      2018) and heavy rainfall events (e.g.    2021)
- Since 1864, mean temperatures in         Thurgau have increased by about 1.9 °C.
- Due to its importance for fruit          production and its strong exposure to    climate change, Güttingen is well        suited for further studies in this       module

**Identifying nearby weather stations**

```{r}
library(chillR)
station_list<-handle_gsod(action="list_stations",
                          location=c( 8.77, 47.73),
                          time_interval=c(1990,2024))
head(station_list)
```

## Downloading historical weather data

**Download weather data for the most promising station on the list**

```{r}
weather<-handle_gsod(action="download_weather",
                     location=station_list$chillR_code[14],
                     time_interval=c(1990,2024))
head(weather[[1]])
```

## Preparing weather data for chillR

**Convert the weather data into chillR format**

```{r}
cleaned_weather<-handle_gsod(weather)
güttingen_weather <- cleaned_weather[[1]]

dir.create('data_güttingen')

write.csv(station_list,"data_güttingen/station_list.csv",row.names=FALSE)
write.csv(weather[[1]],"data_güttingen/Güttingen_raw_weather.csv",row.names=FALSE)
write.csv(cleaned_weather[[1]],"data_güttingen/Güttingen_chillR_weather.csv",row.names=FALSE)
```

## Sources
Genusserbe Bayern – Lake Constance climate and viticulture
https://www.genusserbe.bayern.de/355028/index.php

Arenenberg Thurgau – Experimental fruit production in Güttingen
https://arenenberg.tg.ch/beratung/versuche/obstbau-guettingen.html/654

Arenenberg Thurgau – Fruit production in Thurgau
https://arenenberg.tg.ch/beratung/obst.html/342

Canton of Thurgau – Climate change in Thurgau
https://klima.tg.ch/klima-im-thurgau.html/16119