# Chapter 23 — PLS regression with agroclimatic metrics

This chapter applies PLS regression with agroclimatic metrics to separate chilling and forcing processes and assess whether their effects on phenology can be clearly distinguished.


## PLS regression using chill and heat metrics

**Repeat the PLS_chill_force procedure for the ‘Roter Boskoop’ dataset. Include plots of daily chill and heat accumulation**

**Load packages and prepare hourly temperature data**

```{r}
library(chillR)
library(tidyverse)

temps_hourly <- read_tab("data/TMaxTMin1958-2019_patched.csv") %>%
  stack_hourly_temps(latitude = 50.6)

head(temps_hourly$hourtemps)
```

**Calculate daily chill and heat metrics**

```{r}
daychill <- daily_chill(hourtemps = temps_hourly,
                        running_mean = 1,
                        models = list(
                          Chilling_Hours = Chilling_Hours,
                          Utah_Chill_Units = Utah_Model,
                          Chill_Portions = Dynamic_Model,
                          GDH = GDH)
)

head(daychill$daily_chill)
```

**Visualize daily chill accumulation**

```{r, warning=FALSE}
dc <- make_daily_chill_plot2(daychill,
                             metrics = c("Chill_Portions"),
                             cumulative = FALSE,
                             startdate = 300,
                             enddate = 30,
                             focusyears = c(2008), 
                             metriclabels = "Chill Portions")

dc <- make_daily_chill_plot2(daychill,
                             metrics = c("Chill_Portions"),
                             cumulative = TRUE,
                             startdate = 300,
                             enddate = 30,
                             focusyears = c(2008),
                             metriclabels = "Chill Portions")
```

**Prepare phenological bloom data (Roter Boskoop)**

```{r}
Roter_first <- read_tab("data/Roter_Boskoop_bloom_1958_2019.csv") %>%
  select(Pheno_year, First_bloom) %>%
  mutate(Year = as.numeric(substr(First_bloom, 1, 4)),
         Month = as.numeric(substr(First_bloom, 5, 6)),
         Day = as.numeric(substr(First_bloom, 7, 8))) %>%
  make_JDay() %>%
  select(Pheno_year, 
         JDay) %>%
  rename(Year = Pheno_year,
         pheno = JDay)
```

**Run PLS regression with chill and heat metrics**

```{r}
plscf <- PLS_chill_force(daily_chill_obj = daychill,
                         bio_data_frame = Roter_first,
                         split_month = 6,
                         chill_models = "Chill_Portions",
                         heat_models = "GDH")

head(plscf$Chill_Portions$GDH$PLS_summary)
```

**Define plotting function for PLS chill–forcing results**

```{r}

plot_PLS_chill_force <- function(plscf,
                                 chill_metric = "Chill_Portions",
                                 heat_metric = "GDH",
                                 chill_label = "CP",
                                 heat_label = "GDH",
                                 chill_phase = c(-48, 62),
                                 heat_phase = c(3, 105.5))
{
  PLS_gg <- plscf[[chill_metric]][[heat_metric]]$PLS_summary %>%
    mutate(Month = trunc(Date/100),
           Day = Date - Month * 100,
           Date = ISOdate(2002,
                          Month,
                          Day))
  
  PLS_gg[PLS_gg$JDay <= 0,"Date"]<-
    ISOdate(2001,
            PLS_gg$Month[PLS_gg$JDay <= 0],
            PLS_gg$Day[PLS_gg$JDay <= 0])
  
  PLS_gg <- PLS_gg %>%
    mutate(VIP_importance = VIP >= 0.8,
           VIP_Coeff = factor(sign(Coef) * VIP_importance))
  
  chill_start_date <- ISOdate(2001,
                              12,
                              31) + chill_phase[1] * 24 * 3600
  chill_end_date <- ISOdate(2001,
                            12,
                            31) + chill_phase[2] * 24 * 3600
  heat_start_date <- ISOdate(2001,
                             12,
                             31) + heat_phase[1] * 24 * 3600
  heat_end_date <- ISOdate(2001,
                           12,
                           31) + heat_phase[2] * 24 * 3600
  
  
  
  
  temp_plot <- ggplot(PLS_gg) +
    annotate("rect",
             xmin = chill_start_date,
             xmax = chill_end_date,
             ymin = -Inf,
             ymax = Inf,
             alpha = .1,
             fill = "blue") +
    annotate("rect",
             xmin = heat_start_date,
             xmax = heat_end_date,
             ymin = -Inf,
             ymax = Inf,
             alpha = .1,
             fill = "red") +
    annotate("rect",
             xmin = ISOdate(2001,
                            12,
                            31) +
               min(plscf$pheno$pheno,
                   na.rm = TRUE) * 24 * 3600,
             xmax = ISOdate(2001,
                            12,
                            31) +
               max(plscf$pheno$pheno,
                   na.rm = TRUE) * 24 * 3600,
             ymin = -Inf,
             ymax = Inf,
             alpha = .1,
             fill = "black") +
    geom_vline(xintercept = ISOdate(2001,
                                    12,
                                    31) +
                 median(plscf$pheno$pheno,
                        na.rm=TRUE) * 24 * 3600,
               linetype = "dashed") +
    geom_ribbon(aes(x = Date,
                    ymin = MetricMean - MetricStdev ,
                    ymax = MetricMean + MetricStdev ),
                fill = "grey") +
    geom_ribbon(aes(x = Date,
                    ymin = MetricMean - MetricStdev * (VIP_Coeff == -1),
                    ymax = MetricMean + MetricStdev * (VIP_Coeff == -1)),
                fill = "red") +
    geom_ribbon(aes(x = Date,
                    ymin = MetricMean - MetricStdev * (VIP_Coeff == 1),
                    ymax = MetricMean + MetricStdev * (VIP_Coeff == 1)),
                fill = "dark green") +
    geom_line(aes(x = Date,
                  y = MetricMean)) +
    facet_wrap(vars(Type),
               scales = "free_y",
               strip.position = "left",
               labeller = 
                 labeller(
                   Type =
                     as_labeller(c(Chill = paste0("Chill (",
                                                  chill_label,
                                                  ")"),
                                   Heat = paste0("Heat (",
                                                 heat_label,
                                                 ")"))))) +
    ggtitle("Daily chill and heat accumulation rates") +
    theme_bw(base_size = 15) + 
    theme(strip.background = element_blank(),
          strip.placement = "outside",
          strip.text.y = element_text(size = 12),
          plot.title = element_text(hjust = 0.5),
          axis.title.y = element_blank()
    )
  
  VIP_plot <- ggplot(PLS_gg,
                     aes(x = Date,
                         y = VIP)) +
    annotate("rect",
             xmin = chill_start_date,
             xmax = chill_end_date,
             ymin = -Inf,
             ymax = Inf,
             alpha = .1,
             fill = "blue") +
    annotate("rect",
             xmin = heat_start_date,
             xmax = heat_end_date,
             ymin = -Inf,
             ymax = Inf,
             alpha = .1,
             fill = "red") +
    annotate("rect",
             xmin = ISOdate(2001,
                            12,
                            31) + min(plscf$pheno$pheno,
                                      na.rm = TRUE) * 24 * 3600,
             xmax = ISOdate(2001,
                            12,
                            31) + max(plscf$pheno$pheno,
                                      na.rm = TRUE) * 24 * 3600,
             ymin = -Inf,
             ymax = Inf,
             alpha = .1,
             fill = "black") +
    geom_vline(xintercept = ISOdate(2001,
                                    12,
                                    31) + median(plscf$pheno$pheno,
                                                 na.rm = TRUE) * 24 * 3600,
               linetype = "dashed") +
    geom_bar(stat = 'identity',
             aes(fill = VIP>0.8)) +
    facet_wrap(vars(Type), 
               scales = "free",
               strip.position = "left",
               labeller = 
                 labeller(
                   Type = as_labeller(c(Chill="VIP for chill",
                                        Heat="VIP for heat")))) +
    scale_y_continuous(
      limits = c(0,
                 max(plscf[[chill_metric]][[heat_metric]]$PLS_summary$VIP))) +
    ggtitle("Variable Importance in the Projection (VIP) scores") +
    theme_bw(base_size = 15) + 
    theme(strip.background = element_blank(),
          strip.placement = "outside",
          strip.text.y = element_text(size = 12),
          plot.title = element_text(hjust = 0.5),
          axis.title.y = element_blank()
    ) +
    scale_fill_manual(name = "VIP", 
                      labels = c("<0.8", ">0.8"), 
                      values = c("FALSE" = "grey",
                                 "TRUE" = "blue")) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_blank())
  
  coeff_plot <- ggplot(PLS_gg,
                       aes(x = Date,
                           y = Coef)) +
    annotate("rect",
             xmin = chill_start_date,
             xmax = chill_end_date,
             ymin = -Inf,
             ymax = Inf,
             alpha = .1,
             fill = "blue") +
    annotate("rect",
             xmin = heat_start_date,
             xmax = heat_end_date,
             ymin = -Inf,
             ymax = Inf,
             alpha = .1,
             fill = "red") +
    annotate("rect",
             xmin = ISOdate(2001,
                            12,
                            31) + min(plscf$pheno$pheno,
                                      na.rm = TRUE) * 24 * 3600,
             xmax = ISOdate(2001,
                            12,
                            31) + max(plscf$pheno$pheno,
                                      na.rm = TRUE) * 24 * 3600,
             ymin = -Inf,
             ymax = Inf,
             alpha = .1,
             fill = "black") +
    geom_vline(xintercept = ISOdate(2001,
                                    12,
                                    31) + median(plscf$pheno$pheno,
                                                 na.rm = TRUE) * 24 * 3600,
               linetype = "dashed") +
    geom_bar(stat = 'identity',
             aes(fill = VIP_Coeff)) +
    facet_wrap(vars(Type),
               scales = "free",
               strip.position = "left",
               labeller =
                 labeller(
                   Type = as_labeller(c(Chill = "MC for chill",
                                        Heat = "MC for heat")))) +
    scale_y_continuous(
      limits = c(min(plscf[[chill_metric]][[heat_metric]]$PLS_summary$Coef),
                 max(plscf[[chill_metric]][[heat_metric]]$PLS_summary$Coef))) +
    ggtitle("Model coefficients (MC)") +
    theme_bw(base_size = 15) + 
    theme(strip.background = element_blank(),
          strip.placement = "outside",
          strip.text.y = element_text(size = 12),
          plot.title = element_text(hjust = 0.5),
          axis.title.y = element_blank()
    ) +
    scale_fill_manual(name = "Effect direction", 
                      labels = c("Advancing",
                                 "Unimportant",
                                 "Delaying"), 
                      values = c("-1" = "red",
                                 "0" = "grey",
                                 "1" = "dark green")) +
    ylab("PLS coefficient") +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_blank())
  
  library(patchwork)
  
  plot <- (VIP_plot +
             coeff_plot +
             temp_plot +
             plot_layout(ncol = 1,
                         guides = "collect")
  ) & theme(legend.position = "right",
            legend.text = element_text(size = 8),
            legend.title = element_text(size = 10),
            axis.title.x = element_blank())
  
  plot
  
}
```

**Re-run analysis with smoothed daily metrics**

```{r}
daychill <- daily_chill(hourtemps = temps_hourly,
                        running_mean = 11,
                        models = list(Chilling_Hours = Chilling_Hours,
                                      Utah_Chill_Units = Utah_Model,
                                      Chill_Portions = Dynamic_Model,
                                      GDH = GDH)
)

plscf <- PLS_chill_force(daily_chill_obj = daychill,
                         bio_data_frame = Roter_first,
                         split_month = 6,
                         chill_models = c("Chilling_Hours",
                                          "Utah_Chill_Units",
                                          "Chill_Portions"),
                         heat_models = c("GDH"))

```

**Plot results for individual chill models**

```{r}
plot_PLS_chill_force(plscf,
                     chill_metric = "Chilling_Hours",
                     heat_metric = "GDH",
                     chill_label = "CH",
                     heat_label = "GDH",
                     chill_phase = c(0,0),
                     heat_phase = c(0,0))

plot_PLS_chill_force(plscf,
                     chill_metric = "Utah_Chill_Units",
                     heat_metric = "GDH",
                     chill_label = "CU",
                     heat_label = "GDH",
                     chill_phase = c(0,0),
                     heat_phase = c(0,0))

plot_PLS_chill_force(plscf)

```

## Delineating chilling and forcing phases across chill models

**Run PLS_chill_force analyses for all three major chill models. Delineate your best estimates of chilling and forcing phases for all of them.**

**Chilling phase:**

- Winter VIP values were intermittent     and rarely exceeded the importance      threshold for extended period
- Model coefficients during winter        alternated between positive and         negative values
- This indicates that winter temperature   effects on bloom timing were            inconsistent
- No clear or continuous chilling phase   could be delineated for any chill       model
- Chilling requirements are likely        fulfilled every year and therefore do   not drive interannual variation in      bloom dates

**Forcing phase:**

- VIP values were consistently high (>    0.8) from late winter to spring
- Model coefficients during this period   were predominantly negative
- Warmer temperatures during this phase   consistently advanced bloom
-> A clear forcing phase was identified

**Best estimates**
Chilling phase: no clearly delineable period

Forcing phase: approximately January to April

## Visualization and interpretation of chilling and forcing phases

**Plot results for all three analyses, including shaded plot areas for the chilling and forcing periods you estimated.**

- Forcing periods were shaded in red      (January–April)
- No chilling period was shaded, as no    consistent chilling phase could be      identified
- The plots visually confirm that bloom   timing is controlled primarily by       spring forcing temperatures